Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples utilisables par Isabelle et Coq'' (CUIC). }

### 2016-01-25 ijcar2016/body.tex
more edits

%%% BEGIN Times font
\usepackage{mathptmx}
\usepackage[scaled=.82]{beramono}
\usepackage[scaled=.86]{helvet}
\DeclareSymbolFont{letters}{OML}{txmi}{m}{it}

%%% END Times font

@@

@@

@@

\author {Andrew Reynolds\inst{1} \and Jasmin Christian Blanchette\inst{2,3} \and \\ Simon~Cruanes\inst{2} \and Cesare~Tinelli \inst{1}}
\authorrunning {A. Reynolds, J. C. Blanchette, C. Tinelli}
\institute{
Department of Computer Science, The University of Iowa, USA
\and
Inria Nancy -- Grand Est \& LORIA, Villers-l\`es-Nancy, France
\and
Max-Planck-Institut f\"ur Informatik, Saarbr\"ucken, Germany
%\and
%\'Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Switzerland
}

@@

We define a translation for a class of
formulas involving admissible recursive function definitions
(Section~\ref{sec:encoding}). A
recursive definition $\forall x : \tau.\;\, \const{f}(x) \teq \mathit{rhs}$
is translated to
$\forall a :
\fargtype{\tau}.\;\, \const{f}(\farg{\con{f}}(a)) \teq \mathit{rhs}[\farg{\con{f}}(a)/x]$, where
$\fargtype{\tau}$ is an uninterpreted \relax{abstract} type and $\farg{\con{f}} :
\fargtype{\tau} \to \tau$ converts the abstract type to the concrete
type. Additional constraints ensure that the abstract values that are relevant
to the formula's satisfiability
exist. The translation preserves satisfiability and, for admissible definitions,
unsatisfiability, and makes finite model finding possible for problems in this
class.\begin{conf} Detailed proofs of correctness are included in a technical
report \cite{our-report}.\end{conf}

@@

Our setting is a monomorphic (or many-sorted) first-order logic
like the one defined by SMT-LIB \cite{smtlib25}.
A \emph{signature} $\Sigma$ consists of
a set $\stypes{\Sigma}$ of %ct type symbols
first-order types
%
%\ct{What the rationale for using the letter $y$ in $\stypes{\Sigma}$?}
%\jb{y as in t\textbf{y}pe. changed to `ty'}
(or sorts) and a set $\sfuns{\Sigma}$ of function symbols over these types.
We assume that signatures always contain a Boolean type $\Bool$ and constants
$\ltrue, \lfalse : \Bool$ for truth and falsity,
an infix equality predicate ${\teq} : \tau \times \tau \to \Bool$
for each $\tau \in \stypes{\Sigma}$,
standard Boolean connectives ($\neg$, $\wedge$, $\vee$, etc.),
and an if--then--else function symbol
$\lite : \Bool \times \tau \times \tau \rightarrow \tau$
for each $\tau \in \stypes{\Sigma}$.
%
We fix an infinite set $\svars{\Sigma}_\tau$ of \emph{variables of type $\tau$}
for each $\tau \in \stypes{\Sigma}$ and
define $\svars{\Sigma}$ as $\bigcup_{\tau \in \stypes{\Sigma}} \svars{\Sigma}_\tau$.
$\Sigma$-terms are built as usual over functions symbols in $\Sigma$ and variables in $\svars{\Sigma}$.
%
Formulas are terms of type $\Bool$.
We write $t\typ{\tau}$ to denote %ct a metavariable ranging over 
terms of type~$\tau$ and $\terms( t )$ to denote the set of subterms in $t$.
Given a term $u$, we write $u[\vec t/\vec x]$ to denote the result of replacing
all occurrences of $\vec x$ with $\vec t$ in $u$.
\begin{rep}When applied to terms, the symbol $=$ denotes syntactic equality.\end{rep}

@@

\begin{definition}\rm
A formula $\varphi$ is in \emph{definitional form with respect to}
$\{ \con{f}_1, \ldots, \con{f}_n \} \subseteq \sfuns{\Sigma}$ if it is of the
form
%
%\kern-\abovedisplayskip
%\kern+\abovedisplayshortskip
%
$(\forallf{\con{f}_1} \vec x_1.\; \varphi_1) \wedge \cdots \wedge
(\forallf{\con{f}_n} \vec x_n.\; \varphi_n) \wedge \psi$,
%
where $\con{f}_1, \ldots, \con{f}_n$ are distinct function symbols,
%ct added
$\forallf{\con{f}_i} \vec x_i.\; \varphi_i$ is a function definition
for $i = 1, \ldots, n$,
%
%\ct{I think we need more conditions here on how a symbol $\con{f}_i$ can occur in $\varphi_j$.}
%\jb{do we? I thought that the admissibility criterion below is enough.}
% should be enough, unless I'm missing something? -ajr
and $\psi$ contains no function definitions.
We call $\psi$ the \emph{goal} of $\varphi$.
%\ct{Don't we need additional restrictions on the $\varphi_i$'s here?
%(E.g., they are quantifier-free, or perhaps existential or containing only let binders.)}
% Now fixed above -ajr
\end{definition}

@@

First-order logic has no built-in notion of computation or termination. To ensure
that a function specification is well founded, it is sufficient to require that
the function would terminate when seen as a functional program, under \emph{some}
evaluation order. For example, the definition
$\forall x : {\typeint}.\;\, \lite\bigl(
x \leq 0,\allowbreak\; \con{p}( x ) \teq 1,\allowbreak\; \con{p}( x ) \teq 2 * \con{p}( x - 1 ) \bigr)$,
where $T$ is integer arithmetic extended with the uninterpreted symbol 
$\con p:\typeint \rightarrow \typeint$, can be shown well founded under a strategy that
evaluates the condition of an $\lite$ before evaluating the relevant branch,
ignoring the other branch. \begin{rep}Logically, such dependencies can be captured by
congruence rules. \end{rep}Krauss developed these ideas in the more general context of
higher-order logic \cite[Section 2]{krauss-2009-phd}.

\begin{theorem}\label{thm:adm}
If $\DDD$ is a set of well-founded function definitions for\/
$\sfundefs{\Sigma} \begin{rep} = \{\con{f}_1,\ldots,\con{f}_n\}\end{rep}$, then it is admissible.
\end{theorem}
\begin{rep}
\begin{proofsketch}
Let $\psi$ be a satisfiable formula closed under function expansion with
respect to $\DDD$. We show that $\psi \wedge \bigwedge \Delta$ is also
satisfiable. Let $\I$ be a model of $\psi$, and
let $\I_0$ be the restriction of $\I$ to the function symbols in $\sfuns{\Sigma} -
\sfundefs{\Sigma}$. Because well-founded definitions uniquely characterize
the interpretation of the functions they define, there exists a
$\Sigma$-interpretation $\J$ that extends $\I_0$ such that $\J \models \Delta$.
%
Since $\psi$ is closed under function expansion, it already constrains the
functions in $\sfundefs{\Sigma}$ recursively as far as is
necessary for interpreting $\psi$. Thus, any point $v$ for which
$\const{f}_i^\I(v)$ is needed for interpreting $\psi$ will have its
expected value according to its definition and hence coincide with $\J$.
And since $\psi^\I$ does not depend on the interpretation at the other
points, $\J$ is, like $\I$, a model of $\psi$.
Since $\J \models \Delta$ by assumption, we have $\J \models \psi \wedge
\bigwedge \Delta$ as desired.
%We have
%$\varphi \,\models_T\, \bigwedge\nolimits_{\,i=1}^{n} \{ \varphi_i[ \vec t / \vec x ] \mid \con{f}_i( \vec t ) \in \terms( \varphi ) \}$.
\qed
\end{proofsketch}
\end{rep}

TODO: Say something about admissibility for tail-recursion: Watch out if the
argument always changes...

\begin{rep}
Another useful fragment of function definitions is the class of
\emph{productive} corecursive functions. Corecursive functions are functions to
a coalgebraic datatype. These functions can be ill founded, without
their being inconsistent. Productive corecursive functions are functions that
progressively reveal parts of their potentially infinite
output \cite{turner-1995,mcbride-productive}.
For example, given a type of infinite streams constructed by
$\con{scons} : \ty{int} \times \ty{stream} \to \ty{stream}$,
the definition
$\forallf{\const{e}} x.\;\, \const{e}(x) \teq \con{scons}(x,\; \const{e}(x + 1))$
falls within this fragment: Each call to $\const{e}$ produces one
constructor before entering the nested call. Like terminating recursion,
productive corecursion totally specify the functions it defines.
It is even possible to mix recursion and corecursion in the same function
\cite{blanchette-et-al-2015-fouco}. Theorem~\ref{thm:adm} can be extended to
cover such specifications, based on the observation that unfolding
a corecursive definition infinitely computes a unique infinite object.

Beyond totality, an admissible set can contain underspecified functions
such as $\forallf{\con{f}} x : \typeint.\allowbreak\;\, \con{f}( x )
\teq \con{f}( x - 1 )$ or $\forallf{\con{g}} x.\allowbreak\;\, \con{g}( x
) \teq \con{g}( x )$. We conjecture that one can ignore all
tail-recursive calls (i.e., calls that occupy the right-hand side of the
definition, potentially under some $\lite$ branch) when establishing well-foundedness
or productivity, without affecting admissibility.
\end{rep}

@@

\begin{figure}[t]
\normalsize
\begin{enumerate}
%\begin{framed}
\itemx[\ ]
$\conv_0( t\typ{\tau},\, p )$ $=$
%\\[-.8\baselineskip]
 \begin{itemize}
   \itemx[] $\mathsf{if}$ $\tau = \Bool$ $\mathsf{and}$ $t = \boolop(t_1,\ldots,t_n)$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( t'_i,\, \chi_i ) = \conv_0( t_i,\, \pol( \boolop,\, i,\, p ) )$ $\mathsf{for}$ $i = 1, \ldots, n$ $\mathsf{in}$%\kern6.35em$\mid$
      \itemx[] $\mathsf{let}$ $\chi = \chi_1 \wedge \cdots \wedge \chi_n$ $\mathsf{in}$
      \itemx[] $\mathsf{if}$ $p = \ppos$ $\mathsf{then}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n) ) \wedge \chi,\, \ltrue$}
      \itemx[] $\mathsf{else}$ $\mathsf{if}$ $p = \pneg$ $\mathsf{then}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n) \vee \neg \chi,\, \ltrue$}
      \itemx[] $\mathsf{else}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n),\, \chi$}
%\\[-.8\baselineskip]
    \end{itemize}
  \itemx[] $\mathsf{else}$ $\mathsf{if}$ $t = \forallf{\con{f}} \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$ \bigtuple{$\forall a : \fargtype{\con{f}}.\; u' [ \vecfarg{\con{f}}( a ) / \vec x ],\, \ltrue$}
%\\[-.8\baselineskip]
    \end{itemize}
  \itemx[] $\mathsf{else}$ $\mathsf{if}$ $t = \forall \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$ \bigtuple{$\forall \vec x.\; u',\, \forall \vec x.\; \chi$}
%\\[-.8\baselineskip]
    \end{itemize}
   \itemx[] $\mathsf{else}$
   \begin{itemize}
     \itemx[] \return \bigtuple{$t,\; \bigwedge\, \{ \exists a : \fargtype{\con{f}}.\; \vecfarg{\con{f}}( a ) \teq \vec s \mid \con{f}( \vec s ) \in \terms( t ),\, \con{f} \in \sfundefs{\Sigma} \}$}
%\\[-.8\baselineskip]
   \end{itemize}
 \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\itemx[\ ]
$\conv( \varphi )$ $=$ $\mathsf{let}$ $( \varphi',\, \chi ) = \conv_0( \varphi,\, \ppos )$ $\mathsf{in}$ $\varphi'$
%\\[-1.25\baselineskip] %% TYPESETTING
\end{enumerate}
\caption{\,Definition of translation $\conv$}
\label{fig:A}
\end{figure}

@@

The translation alters the formula $\varphi$ in two ways. First, it restricts the
quantification on function definitions for $\con{f}$ to the corresponding
uninterpreted type $\fargtype{\con{f}}$, inserting applications of the concretization functions $\fargx{\con{f}}{i}$ as needed. 
Second, it augments $\varphi$ with additional constraints of the form
$\exists a :\nobreak \fargtype{\con{f}}.\;\, \vecfarg{\con{f}}( a ) \teq \vec s$,
where $\vecfarg{\con{f}}( a ) \teq \vec s$ abbreviates the formula
$\bigwedge_{i=1}^n\, \fargx{\con{f}}{i}(a) \teq s_i$
with $\vec s = (s_1,\ldots,s_n)$.
These existential %%% for typesetting
constraints
ensure that the restricted definition for $\con{f}$ covers all relevant tuples
of terms, namely those occurring in applications of $\con{f}$ 
that are relevant to the satisfiability of $\varphi$.

If $t$ is an application of a predicate symbol $\boolop$, including the
operators ${\oldneg}$, ${\wedge}$, ${\vee}$, ${\teq}$, and ${\lite}$,
$\conv_0$ calls itself recursively on the arguments $t_i$ and polarity $\pol(
\boolop, i, p )$, with $\pol$ defined as
\[\pol( \boolop, i, p ) =
\begin{cases}
p & \text{if either $\boolop \in \{{\wedge}, {\vee}\}$ or $\boolop = \lite$ and $i \in \{2, 3\}$} \\[-\jot]
\pnegate p & \text{if $\boolop = {\neg}$} \\[-\jot]
\pnone & \text{otherwise}
\end{cases}\]
where $\pnegate p$ is $\pneg$ if $p$ is $\ppos$, $\ppos$
if $p$ is $\pneg$, and $\pnone$ if $p$ is $\pnone$.
The term $t$ is then reconstructed as $\boolop(t_1',\ldots,t_n')$ 
where each $t_i'$ is the result of the recursive call with argument $t_i$. 
If the polarity $p$ associated with $t$ is $\ppos$, 
$\conv_0$ conjunctively adds to $\boolop(t_1',\ldots,t_n')$ the constraint $\chi$ derived from
the subterms. Dually, if $p$ is $\pneg$, it adds a
disjunction with the negated constraint, to achieve the same overall effect.
%ct  as in the positive case. 
It $p$ is $\pnone$, the constraint $\chi$ is
returned to the caller.
%
\begin{rep}\par\end{rep}
%
If $t$ is a function definition, % of a function in $\sfundefs{\Sigma}$,
$\conv_0$ constructs a quantified formula over a single variable
$a$ of type $\fargtype{\con{f}}$ and replaces all occurrences of $\vec x$ in
the body of that formula with $\vecfarg{\con{f}}( a )$. 
(Since function definitions are top-level conjuncts, %by case analysis on the return values of $\conv$,
$\chi$ must be $\ltrue$ and can be ignored.)
%
If $t$ is an unannotated quantified formula, $\conv_0$ calls itself
on the body with the same polarity. \begin{rep}A quantifier is prefixed to the
quantified formula and constraint returned by the recursive call.\end{rep}
Otherwise, $t$ is either an application of an uninterpreted predicate symbol or a term
of a type other than $\Bool$. Then, the returned constraint is a conjunction of
formulas of the form $\exists a : {\fargtype{\con{f}}}.\;\, \vecfarg{\con{f}}(
a ) \teq \vec s$ for each subterm $\con{f}( \vec s )$ of~$t$ such that $\con{f}
\in \sfundefs{\Sigma}$. Such constraints, when asserted positively, ensure that
some element in the abstract domain $\fargtype{\con f}$ is the preimage of
the argument tuple $\vec s$. % under $\vecfarg{\con{f}}$.

@@

Note that the translation $\conv$ results in formulas whose models
(i.e., satisfying interpretations) are generally different from those of $\varphi$.
%\ct{"Preserves models" means that it does not drop models, not that it adds models.}
One model $\I$ for formula~(\ref{eq:ex-after}) in the above example interprets
$\fargtype{\con{s}}$ as a finite set $\{ u_0, \ldots, u_{14} \}$,
$\farg{\con{s}}$ as a finite map $u_i \mapsto i$ for $i = 0, \ldots, 14$,
$\con{c}$ as $14$,
and $\con{s}$ as the almost constant function
%
%\begin{equation} \label{eq:approx-interp}
\begin{conf}$\badlambda$. \end{conf}%
\begin{rep}\[\badlambda\]\end{rep}%
%\end{equation}
%
In other words, $\const{s}$ is interpreted as a function mapping $x$ to the sum
of all positive integers up to $x$ when $0 \leq x \leq 13$, and $105$
otherwise.
The $\Sigma$-reduct of $\I$ is not a model of the original formula~(\ref{eq:ex-before}),
since $\I$ 
%ct wrongly in what sense?
%wrongly 
interprets $\con{s}( n )$ as $105$ when $n < 0$ or $n > 14$.

>>>>

%%% TODO: mention that admissible does not mean finite model
%%% e.g. ite inside or outside, or custom Bool operators

%%% BEGIN Times font
\usepackage{mathptmx}
\usepackage[scaled=.82]{beramono}
\usepackage[scaled=.86]{helvet}
\DeclareSymbolFont{letters}{OML}{txmi}{m}{it}
%%% END Times font

@@

\newcommand\innerkw[1]{\textrm{#1}}

@@

\let\labelitemi=\labelitemii %% CHEAT!
\makeatletter
\ps@myheadings
\makeatother

@@

\author {Andrew Reynolds\inst{1} \and Jasmin Christian Blanchette\inst{2,3} \and \\ Simon~Cruanes\inst{2} \and Cesare~Tinelli \inst{1}}
\authorrunning {A. Reynolds et al.} % , J. C. Blanchette, S. Cruanes, C. Tinelli}
\institute{
Department of Computer Science, The University of Iowa, USA
\and
Inria Nancy -- Grand Est \& LORIA, Villers-l\`es-Nancy, France
\and
Max-Planck-Institut f\"ur Informatik, Saarbr\"ucken, Germany
%\and
%\'Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Switzerland
}

@@

We define a translation for a class of
formulas involving admissible recursive function definitions
(Section~\ref{sec:encoding}). A
recursive definition $\forall x : \tau.\;\, \const{f}(x) \teq \mathit{rhs}$
is translated to
$\forall a :
\fargtype{\tau}.\;\, \const{f}(\farg{\con{f}}(a)) \teq \mathit{rhs}[\farg{\con{f}}(a)/x]$, where
$\fargtype{\tau}$ is an uninterpreted \relax{abstract} type and $\farg{\con{f}} :
\fargtype{\tau} \to \tau$ converts the abstract type to the concrete
type. Additional constraints ensure that the abstract values that are relevant
to the formula's satisfiability
exist. The translation preserves satisfiability and, for admissible definitions,
unsatisfiability, and makes finite model finding possible for problems in this
class.
%\begin{conf} Detailed proofs of correctness are included in a technical
%report \cite{our-report}.\end{conf}

@@

Our setting is a monomorphic (or many-sorted) first-order logic
like the one defined by SMT-LIB \cite{smtlib25}.
A \emph{signature} $\Sigma$ consists of
a set $\stypes{\Sigma}$ of %ct type symbols
first-order types
%
%\ct{What the rationale for using the letter $y$ in $\stypes{\Sigma}$?}
%\jb{y as in t\textbf{y}pe. changed to `ty'}
(or sorts) and a set $\sfuns{\Sigma}$ of function symbols over these types.
We assume that signatures always contain a Boolean type $\Bool$ and constants
$\ltrue, \lfalse : \Bool$ for truth and falsity,
an infix equality predicate ${\teq} : \tau \times \tau \to \Bool$
for each $\tau \in \stypes{\Sigma}$,
standard Boolean connectives ($\neg$, $\wedge$, $\vee$, etc.),
and an \innerkw{if}--\innerkw{then}--\innerkw{else} function symbol
$\lite : \Bool \times \tau \times \tau \rightarrow \tau$
for each $\tau \in \stypes{\Sigma}$.
%
We fix an infinite set $\svars{\Sigma}_\tau$ of \emph{variables of type $\tau$}
for each $\tau \in \stypes{\Sigma}$ and
define $\svars{\Sigma}$ as $\bigcup_{\tau \in \stypes{\Sigma}} \svars{\Sigma}_\tau$.
$\Sigma$-terms are built as usual over functions symbols in $\Sigma$ and variables in $\svars{\Sigma}$.
%
Formulas are terms of type $\Bool$.
We write $t\typ{\tau}$ to denote %ct a metavariable ranging over 
terms of type~$\tau$ and $\terms( t )$ to denote the set of subterms in $t$.
Given a term $u$, we write $u[\vec t/\vec x]$ to denote the result of replacing
all occurrences of $\vec x$ with $\vec t$ in $u$.
\begin{rep}When applied to terms, the symbol $=$ denotes syntactic equality.\end{rep}

@@

\begin{definition}\rm
A formula $\varphi$ is in \emph{definitional form with respect to}
$\{ \con{f}_1, \ldots, \con{f}_n \} \subseteq \sfuns{\Sigma}$ if it is of the
form
%
%\kern-\abovedisplayskip
%\kern+\abovedisplayshortskip
%
$(\forallf{\con{f}_1} \vec x_1.\; \varphi_1) \wedge \cdots \wedge
(\forallf{\con{f}_n} \vec x_n.\; \varphi_n) \wedge \psi$,
%
where $\con{f}_1, \ldots, \con{f}_n$ are distinct function symbols,
%ct added
$\forallf{\con{f}_i} \vec x_i.\; \varphi_i$ is a function definition
for $i = 1, \ldots, n$,
%
%\ct{I think we need more conditions here on how a symbol $\con{f}_i$ can occur in $\varphi_j$.}
%\jb{do we? I thought that the admissibility criterion below is enough.}
% should be enough, unless I'm missing something? -ajr
and $\psi$ contains no function definitions.
We call $\psi$ the \emph{main formula} of $\varphi$.
%\ct{Don't we need additional restrictions on the $\varphi_i$'s here?
%(E.g., they are quantifier-free, or perhaps existential or containing only let binders.)}
% Now fixed above -ajr
\end{definition}

@@

First-order logic has no built-in notion of computation or termination. To ensure
that a function specification is well founded, it is sufficient to require that
the function would terminate when seen as a functional program, under \emph{some}
evaluation order. For example, the definition
$\forall x : {\typeint}.\;\, 
\con{p}( x ) \teq \lite\bigl( x \leq 0,\allowbreak\;  1,\allowbreak\;
  2 * \con{p}( x - 1 ) \bigr)$,
where $T$ is integer arithmetic extended with the uninterpreted symbol 
$\con p:\typeint \rightarrow \typeint$, can be shown well founded under a strategy that
evaluates the condition of an $\lite$ before evaluating the relevant branch,
ignoring the other branch. Logically, such dependencies can be captured by
congruence rules. Krauss developed these ideas in the more general context of
higher-order logic \cite[Section 2]{krauss-2009-phd}.

\begin{theorem}\label{thm:adm}
If $\DDD$ is a set of well-founded function definitions for\/
$\sfundefs{\Sigma} = \{\con{f}_1,\ldots,\con{f}_n\}$, then it is admissible.
\end{theorem}
\begin{proofsketch}
Let $\psi$ be a satisfiable formula closed under function expansion with
respect to $\DDD$. We show that $\psi \wedge \bigwedge \Delta$ is also
satisfiable. Let $\I$ be a model of $\psi$, and
let $\I_0$ be the restriction of $\I$ to the function symbols in $\sfuns{\Sigma} -
\sfundefs{\Sigma}$. Because well-founded definitions uniquely characterize
the interpretation of the functions they define, there exists a
$\Sigma$-interpretation $\J$ that extends $\I_0$ such that $\J \models \Delta$.
%
Since $\psi$ is closed under function expansion, it already constrains the
functions in $\sfundefs{\Sigma}$ recursively as far as is
necessary for interpreting $\psi$. Thus, any point $v$ for which
$\const{f}_i^\I(v)$ is needed for interpreting $\psi$ will have its
expected value according to its definition and hence coincide with $\J$.
And since $\psi^\I$ does not depend on the interpretation at the other
points, $\J$ is, like $\I$, a model of $\psi$.
Since $\J \models \Delta$ by assumption, we have $\J \models \psi \wedge
\bigwedge \Delta$ as desired.
%We have
%$\varphi \,\models_T\, \bigwedge\nolimits_{\,i=1}^{n} \{ \varphi_i[ \vec t / \vec x ] \mid \con{f}_i( \vec t ) \in \terms( \varphi ) \}$.
\qed
\end{proofsketch}

% TODO: Say something about admissibility for tail-recursion: Watch out if the
% argument always changes...

Another useful fragment of function definitions is the class of
\emph{productive} corecursive functions. Corecursive functions are functions to
a coalgebraic datatype. These functions can be ill founded, without
their being inconsistent. Productive corecursive functions are functions that
progressively reveal parts of their potentially infinite
output \cite{turner-1995,mcbride-productive}.
For example, given a type of infinite streams constructed by
$\con{scons} : \ty{int} \times \ty{stream} \to \ty{stream}$,
the definition
$\forallf{\const{e}} x.\;\, \const{e}(x) \teq \con{scons}(x,\; \const{e}(x + 1))$
falls within this fragment: Each call to $\const{e}$ produces one
constructor before entering the nested call. Like terminating recursion,
productive corecursion totally specify the functions it defines.
It is even possible to mix recursion and corecursion in the same function
\cite{blanchette-et-al-2015-fouco}. Theorem~\ref{thm:adm} can be extended to
cover such specifications, based on the observation that unfolding
a corecursive definition infinitely computes a unique infinite object.
%We conjecture that a large fragment of mixed recursive--corecursive
%specifications are

\begin{rep}
Beyond totality, an admissible set can contain underspecified functions
such as $\forallf{\con{f}} x : \typeint.\allowbreak\;\, \con{f}( x )
\teq \con{f}( x - 1 )$ or $\forallf{\con{g}} x.\allowbreak\;\, \con{g}( x
) \teq \con{g}( x )$. We conjecture that one can ignore all
tail-recursive calls (i.e., calls that occupy the right-hand side of the
definition, potentially under some $\lite$ branch) when establishing well-foundedness
or productivity, without affecting admissibility.
\end{rep}

@@

\begin{figure}[t]
\normalsize
\begin{enumerate}
%\begin{framed}
\itemx[\ ]
$\conv_0( t\typ{\tau},\, p )$ $=$
%\\[-.8\baselineskip]
 \begin{itemize}
   \itemx[] $\innerkw{if}$ $\tau = \Bool$ $\mathsf{and}$ $t = \boolop(t_1,\ldots,t_n)$ $\innerkw{then}$
    \begin{itemize}
      \itemx[] $\innerkw{let}$ $( t'_i,\, \chi_i ) = \conv_0( t_i,\, \pol( \boolop,\, i,\, p ) )$ $\mathsf{for}$ $i = 1, \ldots, n$ $\innerkw{in}$%\kern6.35em$\mid$
      \itemx[] $\innerkw{let}$ $\chi = \chi_1 \wedge \cdots \wedge \chi_n$ $\innerkw{in}$
      \itemx[] $\innerkw{if}$ $p = \ppos$ $\innerkw{then}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n) ) \wedge \chi,\, \ltrue$}
      \itemx[] $\innerkw{else}$ $\innerkw{if}$ $p = \pneg$ $\innerkw{then}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n) \vee \neg \chi,\, \ltrue$}
      \itemx[] $\innerkw{else}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n),\, \chi$}
%\\[-.8\baselineskip]
    \end{itemize}
  \itemx[] $\innerkw{else}$ $\innerkw{if}$ $t = \forallf{\con{f}} \vec x.\;\, u$ $\innerkw{then}$
    \begin{itemize}
      \itemx[] $\innerkw{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\innerkw{in}$ \bigtuple{$\forall a : \fargtype{\con{f}}.\; u' [ \vecfarg{\con{f}}( a ) / \vec x ],\, \ltrue$}
%\\[-.8\baselineskip]
    \end{itemize}
  \itemx[] $\innerkw{else}$ $\innerkw{if}$ $t = \forall \vec x.\;\, u$ $\innerkw{then}$
    \begin{itemize}
      \itemx[] $\innerkw{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\innerkw{in}$ \bigtuple{$\forall \vec x.\; u',\, \forall \vec x.\; \chi$}
%\\[-.8\baselineskip]
    \end{itemize}
   \itemx[] $\innerkw{else}$
   \begin{itemize}
     \itemx[] \return \bigtuple{$t,\; \bigwedge\, \{ \exists a : \fargtype{\con{f}}.\; \vecfarg{\con{f}}( a ) \teq \vec s \mid \con{f}( \vec s ) \in \terms( t ),\, \con{f} \in \sfundefs{\Sigma} \}$}
%\\[-.8\baselineskip]
   \end{itemize}
 \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\itemx[\ ]
$\conv( \varphi )$ $=$ $\innerkw{let}$ $( \varphi',\, \chi ) = \conv_0( \varphi,\, \ppos )$ $\innerkw{in}$ $\varphi'$
\\[-1.5\baselineskip] %% TYPESETTING
\end{enumerate}
\caption{\,Definition of translation $\conv$}
\label{fig:A}
\end{figure}

@@

The translation alters the formula $\varphi$ in two ways. First, it restricts the
quantification on function definitions for $\con{f}$ to the corresponding
uninterpreted type $\fargtype{\con{f}}$, inserting applications of the concretization functions $\fargx{\con{f}}{i}$ as needed. 
Second, it augments $\varphi$ with additional constraints of the form
$\exists a :\nobreak \fargtype{\con{f}}.\;\, \vecfarg{\con{f}}( a ) \teq \vec s$,
where $\vecfarg{\con{f}}( a ) \teq \vec s$ abbreviates the formula
$\bigwedge_{i=1}^n\, \fargx{\con{f}}{i}(a) \teq s_i$
with $\vec s = (s_1,\ldots,s_n)$.
These existential %%% for typesetting
constraints
ensure that the restricted definition for $\con{f}$ covers all relevant tuples
of terms, namely those occurring in applications of $\con{f}$ 
that are relevant to the satisfiability of $\varphi$. The constraints are
generated as deep in the formula as possible, based on our knowledge of the
polarities of logical connectives, to allow models in which the domains
interpreting the $\fargtype{\con{f}}$ types are as small as possible.

If $t$ is an application of a predicate symbol $\boolop$, including the
operators ${\oldneg}$, ${\wedge}$, ${\vee}$, ${\teq}$, and ${\lite}$,
$\conv_0$ calls itself recursively on the arguments $t_i$ and polarity $\pol(
\boolop, i, p )$, with $\pol$ d\rlap{efined as}
\[\pol( \boolop, i, p ) =
\begin{cases}
p & \text{if either $\boolop \in \{{\wedge}, {\vee}\}$ or $\boolop = \lite$ and $i \in \{2, 3\}$} \\[-\jot]
\pnegate p & \text{if $\boolop = {\neg}$} \\[-\jot]
\pnone & \text{otherwise}
\end{cases}\]
where $\pnegate p$ is $\pneg$ if $p$ is $\ppos$, $\ppos$
if $p$ is $\pneg$, and $\pnone$ if $p$ is $\pnone$.
The term $t$ is then reconstructed as $\boolop(t_1',\ldots,t_n')$ 
where each $t_i'$ is the result of the recursive call with argument~$t_i$. 
If the polarity $p$ associated with $t$ is $\ppos$, 
$\conv_0$ conjunctively adds to $\boolop(t_1',\ldots,t_n')$ the constraint $\chi$ derived from
the subterms and returns $\top$ as the constraint.
Dually, if $p$ is $\pneg$, it adds a
disjunction with the negated constraint, to achieve the same overall effect.
%ct  as in the positive case. 
It $p$ is $\pnone$, the constraint $\chi$ is
returned to the caller.

If $t$ is a function definition, % of a function in $\sfundefs{\Sigma}$,
$\conv_0$ constructs a quantified formula over a single variable
$a$ of type $\fargtype{\con{f}}$ and replaces all occurrences of $\vec x$ in
the body of that formula with $\vecfarg{\con{f}}( a )$. 
(Since function definitions are top-level conjuncts, %by case analysis on the return values of $\conv$,
$\chi$ must be $\ltrue$ and can be ignored.)
%
If $t$ is an unannotated quantified formula, $\conv_0$ calls itself
on the body with the same polarity; a quantifier is prefixed to the
quantified formula and constraint returned by the recursive call.
Otherwise, $t$ is either an application of an uninterpreted predicate symbol or a term
of a type other than $\Bool$. Then, the returned constraint is a conjunction of
formulas of the form $\exists a : {\fargtype{\con{f}}}.\;\, \vecfarg{\con{f}}(
a ) \teq \vec s$ for each subterm $\con{f}( \vec s )$ of~$t$ such that $\con{f}
\in \sfundefs{\Sigma}$. Such constraints, when asserted positively, ensure that
some element in the abstract domain $\fargtype{\con f}$ is the preimage of
the argument tuple $\vec s$. % under $\vecfarg{\con{f}}$.

@@

Note that the translation $\conv$ results in formulas whose models
(i.e., satisfying interpretations) are generally different from those of $\varphi$.
%\ct{"Preserves models" means that it does not drop models, not that it adds models.}
One model $\I$ for formula~(\ref{eq:ex-after}) in the above example interprets
$\fargtype{\con{s}}$ as a finite set $\{ u_0, \ldots, u_{14} \}$,
$\farg{\con{s}}$ as a finite map $u_i \mapsto i$ for $i = 0, \ldots, 14$,
$\con{c}$ as $14$,
and $\con{s}$ as the almost constant function
%
%\begin{equation} \label{eq:approx-interp}
\[\badlambda\]
%\end{equation}
%
In other words, $\const{s}$ is interpreted as a function mapping $x$ to the sum
of all positive integers up to $x$ when $0 \leq x \leq 13$, and $105$
otherwise.
The $\Sigma$-reduct of $\I$ is not a model of the original formula~(\ref{eq:ex-before}),
since $\I$ 
%ct wrongly in what sense?
%wrongly 
interprets $\con{s}( n )$ as $105$ when $n < 0$ or $n > 14$.
